---
title: 'Quick examples of Rbiips usage'
author: 'Adrien Todeschini, Francois Caron'
date: '`r Sys.Date()`'
output:
  rmarkdown::html_vignette:
    toc: yes
    fig_width: 7
    fig_height: 5
vignette: >
  %\VignetteIndexEntry{Quick examples of Rbiips usage}
  %\VignetteEngine{knitr::rmarkdown}
  %\usepackage[utf8]{inputenc}
---

This vignette gathers the examples used in the Rbiips reference manual.

```{r}
set.seed(0)
require(Rbiips)
```

# Add custom functions and distributions to BUGS language

Add custom function `f`

```{r}
f_dim <- function(x_dim, t_dim) {
  # Check dimensions of the input and return dimension of the output of function f
  stopifnot(prod(x_dim) == 1, prod(t_dim) == 1)
  x_dim
}
f_eval <- function(x, t) {
  # Evaluate function f
  0.5 * x + 25 * x/(1 + x^2) + 8 * cos(1.2 * t)
}
biips_add_function('f', 2, f_dim, f_eval)
```

Add custom sampling distribution `dMN`

```{r}
dMN_dim <- function(mu_dim, Sig_dim) {
  # Check dimensions of the input and return dimension of the output of
  # distribution dMN
  stopifnot(prod(mu_dim) == mu_dim[1], length(Sig_dim) == 2, mu_dim[1] == Sig_dim)
  mu_dim
}
dMN_sample <- function(mu, Sig) {
  # Draw a sample of distribution dMN
  mu + t(chol(Sig)) %*% rnorm(length(mu))
}
biips_add_distribution('dMN', 2, dMN_dim, dMN_sample)
```

# Compile model

```{r}
modelfile <- system.file('extdata', 'hmm_f.bug', package = 'Rbiips')
stopifnot(nchar(modelfile) > 0)
cat(readLines(modelfile), sep = '\n')

data <- list(tmax = 10, p = c(.5, .5), logtau_true = log(1), logtau = log(1))
model <- biips_model(modelfile, data, sample_data = TRUE)

rm(model)

tmax <- 10
p <- c(.5, .5)
logtau_true <- log(1)
logtau <- logtau_true

datanames <- c('tmax', 'p', 'logtau_true', 'logtau')
model <- biips_model(modelfile, datanames, sample_data = TRUE)

is.biips(model)
print(model)

model$data()

variable.names(model)
biips_variable_names(model)

biips_nodes(model)

dotfile <- 'hmm.dot'
biips_print_dot(model, dotfile)
# cat(readLines(dotfile), sep = '\n')
```

# SMC algorithm

```{r}
biips_build_sampler(model, proposal = 'prior')
biips_nodes(model, type = 'stoch', observed = FALSE)

biips_build_sampler(model, proposal = 'auto')
biips_nodes(model, type = 'stoch', observed = FALSE)

n_part <- 100
out_smc <- biips_smc_samples(model, c('x', 'c[2:10]'), n_part, type = 'fs', 
                             rs_thres = 0.5, rs_type = 'stratified')

```

Manipulate `smcarray.fsb.list` object

```{r}
is.smcarray.fsb.list(out_smc)
names(out_smc)
out_smc
biips_diagnosis(out_smc)
biips_summary(out_smc)
```

Manipulate `smcarray.fsb` object

```{r}
is.smcarray.fsb(out_smc$x)
names(out_smc$x)
out_smc$x
biips_diagnosis(out_smc$x)
summ_smc_x <- biips_summary(out_smc$x, order = 2, probs = c(.025, .975))
summ_smc_x
dens_smc_x <- biips_density(out_smc$x, bw = 'nrd0', adjust = 1, n = 100)
par(mfrow = c(2, 2))
plot(dens_smc_x)

is.smcarray.fsb(out_smc[['c[2:10]']])
names(out_smc[['c[2:10]']])
out_smc[['c[2:10]']]
biips_diagnosis(out_smc[['c[2:10]']])
summ_smc_c <- biips_summary(out_smc[['c[2:10]']])
summ_smc_c
table_smc_c <- biips_table(out_smc[['c[2:10]']])
par(mfrow = c(2, 2))
plot(table_smc_c)
```

Manipulate `smcarray` object

```{r}
is.smcarray(out_smc$x$f)
names(out_smc$x$f)
out_smc$x$f
out_smc$x$s
biips_diagnosis(out_smc$x$f)
biips_diagnosis(out_smc$x$s)
biips_summary(out_smc$x$f)
biips_summary(out_smc$x$s)
par(mfrow = c(2, 2))
plot(biips_density(out_smc$x$f))
par(mfrow = c(2, 2))
plot(biips_density(out_smc$x$s))

par(mfrow = c(2, 2))
plot(biips_table(out_smc[['c[2:10]']]$f))
par(mfrow = c(2, 2))
plot(biips_table(out_smc[['c[2:10]']]$s))
```


```{r}
par(mfrow = c(2, 2))
plot(model$data()$x_true, type = 'l', col = 'green', xlab = 't', ylab = 'x[t]')
lines(summ_smc_x$f$mean, col = 'blue')
lines(summ_smc_x$s$mean, col = 'red')
matlines(matrix(unlist(summ_smc_x$f$quant), data$tmax), lty = 2, col = 'blue')
matlines(matrix(unlist(summ_smc_x$s$quant), data$tmax), lty = 2, col = 'red')
legend('topright', leg = c('true', 'SMC filtering estimate', 'SMC smoothing estimate'),
  lty = 1, col = c('green', 'blue', 'red'), bty = 'n')

barplot(.5*(model$data()$c_true==1), col = 'green', border = NA, space = 0, offset=2, 
        ylim=c(0,3), xlab='t', ylab='c[t]==1', axes = FALSE)
axis(1, at=1:data$tmax-.5, labels=1:data$tmax)
axis(2, line = 1, at=c(0,3), labels=NA)
text(data$tmax/2, 2.75, 'true')
barplot(.5*c(NA, summ_smc_c$f$mode==1), col = 'blue', border = NA, space = 0, 
        offset=1, axes = FALSE, add = TRUE)
text(data$tmax/2, 1.75, 'SMC filtering mode')
barplot(.5*c(NA, summ_smc_c$s$mode==1), col = 'red', border = NA, space = 0, 
        axes = FALSE, add = TRUE)
text(data$tmax/2, .75, 'SMC smoothing mode')

t <- 5
plot(dens_smc_x[[t]], col = c('blue','red'), ylab = 'posterior density')
points(model$data()$x_true[t], 0, pch = 17, col = 'green')

plot(table_smc_c[[t-1]], col = c('blue','red'), ylab = 'posterior probability mass')
points(model$data()$c_true[t], 0, pch = 17, col = 'green')
```

# PIMH algorithm

```{r}
modelfile <- system.file('extdata', 'hmm.bug', package = 'Rbiips')
stopifnot(nchar(modelfile) > 0)
cat(readLines(modelfile), sep = '\n')

data <- list(tmax = 10, p = c(.5, .5), logtau_true = log(1), logtau = log(1))
model <- biips_model(modelfile, data)

n_part <- 50
obj_pimh <- biips_pimh_init(model, c('x', 'c[2:10]'))  # Initialize
is.pimh(obj_pimh)
out_pimh_burn <- biips_pimh_update(obj_pimh, 100, n_part)  # Burn-in
out_pimh <- biips_pimh_samples(obj_pimh, 100, n_part)  # Samples
```

Manipulate `mcmcarray.list` object

```{r}
is.mcmcarray.list(out_pimh)
names(out_pimh)
out_pimh
biips_summary(out_pimh)
```

Manipulate `mcmcarray` object

```{r}
is.mcmcarray(out_pimh$x)
out_pimh$x
summ_pimh_x <- biips_summary(out_pimh$x, order = 2, probs = c(0.025, 0.975))
summ_pimh_x
dens_pimh_x <- biips_density(out_pimh$x)
par(mfrow = c(2, 2))
plot(dens_pimh_x)
par(mfrow = c(2, 2))
biips_hist(out_pimh$x)

is.mcmcarray(out_pimh[['c[2:10]']])
out_pimh[['c[2:10]']]
summ_pimh_c <- biips_summary(out_pimh[['c[2:10]']])
summ_pimh_c
table_pimh_c <- biips_table(out_pimh[['c[2:10]']])
par(mfrow = c(2, 2))
plot(table_pimh_c)
```

```{r}
par(mfrow = c(2, 1))
plot(c(out_pimh_burn$log_marg_like, out_pimh$log_marg_like), type = 'l', col = 'blue',
  xlab = 'PIMH iteration', ylab = 'log p(y)')

t <- 5
plot(out_pimh$x[t, ], type = 'l', col = 'blue', xlab = 'PIMH iteration', 
     ylab = paste0('x[',t,']'))
points(0, model$data()$x_true[t], pch = 17, col = 'green')

par(mfrow = c(2, 2))
plot(model$data()$x_true, type = 'l', col = 'green', xlab = 't', ylab = 'x[t]')
lines(summ_pimh_x$mean, col = 'blue')
matlines(matrix(unlist(summ_pimh_x$quant), data$tmax), lty = 2, col = 'blue')
legend('topright', leg = c('true', 'PIMH estimate'), lty = c(2, 1), col = c('green',
  'blue'), bty = 'n')

barplot(.5*(model$data()$c_true==1), col = 'green', border = NA, space = 0, offset = 1, 
        ylim=c(0,2), xlab='t', ylab='c[t]==1', axes = FALSE)
axis(1, at=1:data$tmax-.5, labels=1:data$tmax)
axis(2, line = 1, at=c(0,2), labels=NA)
text(data$tmax/2, 1.75, 'true')
barplot(.5*c(NA, summ_pimh_c$mode==1), col = 'blue', border = NA, space = 0, axes = FALSE, add = TRUE)
text(data$tmax/2, .75, 'PIMH mode')

plot(dens_pimh_x[[t]], col='blue', main = , ylab = 'posterior density')
points(model$data()$x_true[t], 0, pch = 17, col = 'green')

plot(table_pimh_c[[t-1]], col='blue', ylab = 'posterior probability mass')
points(model$data()$c_true[t], 0, pch = 17, col = 'green')
```

# SMC sensitivity analysis

```{r}
modelfile <- system.file('extdata', 'hmm.bug', package = 'Rbiips')
stopifnot(nchar(modelfile) > 0)
cat(readLines(modelfile), sep = '\n')

data <- list(tmax = 10, p = c(.5, .5), logtau_true = log(1), logtau = log(1))
model <- biips_model(modelfile, data)

n_part <- 50
logtau_val <- -10:10
out_sens <- biips_smc_sensitivity(model, list(logtau = logtau_val), n_part)

par(mfrow = c(2, 1))
plot(logtau_val, out_sens$log_marg_like, type = 'l', col = 'blue', 
     xlab = 'logtau', ylab = 'log p(y|logtau) ', main = 'SMC sensitivity')
points(data$logtau, min(out_sens$log_marg_like), pch = 17, col = 'green')

plot(logtau_val, out_sens$log_marg_like_pen, type = 'l', col = 'blue', 
     xlab = 'logtau', ylab = 'log p(y|logtau) + log p(logtau)')
plml <- out_sens$log_marg_like_pen
ymin <- min(plml[is.finite(plml)])
points(data$logtau, ymin, pch = 17, col = 'green')
```

# PMMH algorithm

```{r}
modelfile <- system.file('extdata', 'hmm.bug', package = 'Rbiips')
stopifnot(nchar(modelfile) > 0)
cat(readLines(modelfile), sep = '\n')

data <- list(tmax = 10, p = c(.5, .5), logtau_true = log(1))
model <- biips_model(modelfile, data)

n_part <- 50
obj_pmmh <- biips_pmmh_init(model, 'logtau', latent_names = c('x', 'c[2:10]'), 
                            inits = list(logtau = -2))  # Initialize
is.pmmh(obj_pmmh)
out_pmmh_burn <- biips_pmmh_update(obj_pmmh, 100, n_part)  # Burn-in
out_pmmh <- biips_pmmh_samples(obj_pmmh, 100, n_part, thin = 1)  # Samples
```

Manipulate `mcmcarray.list` object

```{r}
is.mcmcarray.list(out_pmmh)
names(out_pmmh)
out_pmmh
biips_summary(out_pmmh)
```

Manipulate `mcmcarray` object

```{r}
is.mcmcarray(out_pmmh$logtau)
out_pmmh$logtau
summ_pmmh_lt <- biips_summary(out_pmmh$logtau, order = 2, probs = c(0.025, 0.975))
dens_pmmh_lt <- biips_density(out_pmmh$logtau)
par(mfrow = c(2, 1))
plot(dens_pmmh_lt)
biips_hist(out_pmmh$logtau)

is.mcmcarray(out_pmmh$x)
out_pmmh$x
summ_pmmh_x <- biips_summary(out_pmmh$x, order = 2, probs = c(0.025, 0.975))
dens_pmmh_x <- biips_density(out_pmmh$x)
par(mfrow = c(2, 2))
plot(dens_pmmh_x)
par(mfrow = c(2, 2))
biips_hist(out_pmmh$x)

is.mcmcarray(out_pmmh[['c[2:10]']])
out_pmmh[['c[2:10]']]
summ_pmmh_c <- biips_summary(out_pmmh[['c[2:10]']])
table_pmmh_c <- biips_table(out_pmmh[['c[2:10]']])
par(mfrow = c(2, 2))
plot(table_pmmh_c)
```

```{r}
par(mfrow = c(2, 2))
plot(c(out_pmmh_burn$log_marg_like_pen, out_pmmh$log_marg_like_pen), type = 'l',
  col = 'blue', xlab = 'PMMH iteration', ylab = 'log p(y|logtau) + log p(logtau)')

plot(out_pmmh$logtau[1, ], type = 'l', col = 'blue', 
     xlab = 'PMMH iteration', ylab = 'logtau')
points(0, model$data()$logtau_true, pch = 17, col = 'green')

plot(dens_pmmh_lt, col = 'blue', ylab = 'posterior density')
points(model$data()$logtau_true, 0, pch = 17, col = 'green')

biips_hist(out_pmmh$logtau, col = 'blue', ylab = 'posterior density')
points(model$data()$logtau_true, 0, pch = 17, col = 'green')

par(mfrow = c(2, 2))
plot(model$data()$x_true, type = 'l', col = 'green', xlab = 't', ylab = 'x[t]')
lines(summ_pmmh_x$mean, col = 'blue')
matlines(matrix(unlist(summ_pmmh_x$quant), data$tmax), lty = 2, col = 'blue')
legend('topright', leg = c('true', 'PMMH estimate'), lty = c(2, 1), 
       col = c('green', 'blue'), bty = 'n')

barplot(.5*(model$data()$c_true==1), col = 'green', border = NA, space = 0, offset = 1, 
        ylim=c(0,2), xlab='t', ylab='c[t]==1', axes = FALSE)
axis(1, at=1:data$tmax-.5, labels=1:data$tmax)
axis(2, line = 1, at=c(0,2), labels=NA)
text(data$tmax/2, 1.75, 'true')
barplot(.5*c(NA, summ_pmmh_c$mode==1), col = 'blue', border = NA, space = 0, axes = FALSE, add = TRUE)
text(data$tmax/2, .75, 'PMMH mode')

t <- 5
plot(dens_pmmh_x[[t]], col='blue', ylab = 'posterior density')
points(model$data()$x_true[t], 0, pch = 17, col = 'green')

plot(table_pmmh_c[[t-1]], col='blue', ylab = 'posterior probability mass')
points(model$data()$c_true[t], 0, pch = 17, col = 'green')
```
